Fourier-domain modulations and delays of gravitational-wave signals

### 2018-06-17 docs/paper_method.tex
Edits to the paper.

The \jgb{constellation} response~\eqref{eq:GslrL}-\eqref{eq:transferLlocal} can be interpreted as the Fourier-domain translation of a discrete derivative taken on the waveform. The \jgb{leading factor in~\eqref{eq:GslrL}} shows that the amplitude of the response is proportional to $f$ in the low-frequency limit $f\ll f_{L}$, where the other factors are essentially unity. For $f\gtrsim f_{L}$, the $\sinc$ and the phase of the exponential generate additional structure in the response, including zero-crossings when the projected armlength \jgb{is an integer number of wavelengths}. From~\eqref{eq:transferLlocal}, an expansion for small $f\ll f_{L}$ yields back a Fourier-domain analog of the low-frequency approximation of the response~\cite{Cutler97, RCP04}, which \jgb{is equivalent to} having two LIGO-type interferometers turned by $\pi/4$ and set in motion. See Fig.~\ref{} and ~\ref{} \SM{[R]} for examples of LISA transfer functions.

\jgb{For analysis of the response we need a concrete set of gravitational waveforms. We will} use the PhenomD model~\cite{Khan+15,Husa+15}, which provides Fourier-domain inspiral-merger-ringdown waveforms for aligned spins.

@@

To obtain an order-of-magnitude estimate for the error measures $\epsilon$~\eqref{eq:deffom}, we will use the Newtonian-order expressions~\eqref{eq:timescalesN} for the signal-dependent timescales $\Tf$ and $T_{A1}$. Higher-order amplitude terms beyond the first one in~\eqref{eq:resulttaylorA}, as well as phase terms beyond the second derivative in~\eqref{eq:expandPsi}, will turn out to be completely negligible in the LISA case and we will ignore them in the following. It is useful to separate the orbital response~\eqref{eq:transfer0local} and the constellation response~\eqref{eq:transferLlocal}, as the different baseline of the delays (orbital radius $R$ or armlength $L$) as well as the presence of a time-varying prefactor $F(t)$ both affect the result. When estimating the magnitude of the relevant derivatives of $G$, we must also take into account dimensionless delay factors of the type $2\pi f d$.

We start with the orbital response, which takes the form of a pure delay $G_{0}(f, t) = e^{-2 i \pi f d_{0}(t)}$ \jgb{\bf[Why subscripted with zero not O or R?]}, and obtain
\begin{subequations}
\begin{align}
	\frac{1}{G_{0}} \partial_{t} G_{0} &= -2i\pi f \dot{d}_{0}\,,\\
	\frac{1}{G_{0}} \partial_{tt} G_{0} &= -2i\pi f \ddot{d}_{0} - 4\pi^{2} f^{2} \dot{d}_{0}^{2} \,,\\
	\frac{1}{G_{0}} \partial_{tf} G_{0} &= -2 i \pi \dot{d}_{0} - 4\pi^{2} f d_{0} \dot{d}_{0} \,.
\end{align}
\end{subequations}

Since the one-arm constellation response~\eqref{eq:decomposeGslr}-\eqref{eq:transferLlocal} is analogous to a discrete time derivative of the signal, it is appropriate to keep explicit an overall factor $f$ reflecting this structure. Hence we write symbolically $G_{L}(f,t) \sim f F(t) e^{-2i \pi f d_{L}(t)}$, where $d_{L}$ represents a delay term and $F(t)$ represents the rest of the geometric factors in~\eqref{eq:transferLlocal}, \jgb{for which we momentarily} ignore the $f$-dependence. This gives
\begin{subequations}
\begin{align}
	\frac{1}{G_{L}} \partial_{t} G_{L} &\sim -2i\pi f \dot{d}_{c} + \frac{\dot{F}}{F}\,,\\
	\frac{1}{G_{L}} \partial_{tt} G_{L} &\sim -2i\pi f \ddot{d}_{c} - 4\pi^{2} f^{2} \dot{d}_{c}^{2} - 2i\pi f \dot{d}_{c} \frac{\dot{F}}{F} - \frac{\dot{F}^{2}}{F^{2}} + \frac{\ddot{F}}{F} \,,\\
	\frac{1}{G_{L}} \partial_{tf} G_{L} &\sim -4 i \pi \dot{d}_{c} - 2i\pi d_{c} \frac{\dot{F}}{F} - 4\pi^{2} f d_{c} \dot{d}_{c} + \frac{1}{f}\frac{\dot{F}}{F}\,.
\end{align}
\end{subequations}

@@

\jgb{\bf[This paragraph is confusing.  Can we point concretely to these factors somewhere, (e.g. in the expansion of the exponential.xxx)? Also it seems to say that we can't do what we just did. Should discuss.]} An important point in the above is the presence of delay factors in the forms of powers of $2\pi f d$. Thus, the magnitude of the corrections cannot be evaluated with the simple replacement $\partial_{t}^{n} \rightarrow \Omega_{0}^{n}$, and the frequency dependency of the error measures $\epsilon$ will depend on the frequency being above or below the transfer frequencies $f_{R}$ and $f_{L}$ defined in~\eqref{eq:transferfrequencies}. The error measure $\epsilon_{d}$ does not depend on signal-dependent timescales and can be directly read off the estimates above, with $\epsilon_{d} = |\partial_{tf}G/(2\pi G)|$. For the error measures $\epsilon_{\Psi 2}$ and $\epsilon_{A1}$, we can combine the above derivatives with the Newtonian timescales given in~\eqref{eq:timescalesN} for the inspiral phase of the signal.

The starting frequency of the signal will play an important role. As a function of the time remaining before merger $\Delta t$, from the Newtonian relation~\eqref{eq:omegaphiN} we have (using the chirp mass $\Mchirp = M \nu^{3/5}$):
\be\label{eq:fstartN}
	f_{\rm start} = 1.75\times 10^{-5} \Hz \left( \frac{\Mchirp}{10^{6}M_{\odot}} \right)^{-5/8} \left( \frac{\Delta t}{10 \yr} \right)^{-3/8} \,.
\ee
This gives in turn for the Newtonian timescales~\eqref{eq:timescalesN}
\begin{subequations}\label{eq:timescalesNfstart}
\begin{align}
	\Tf^{\rm N} &= 8.78\times10^{-2}\yr \left( \frac{\Mchirp}{10^{6}M_{\odot}} \right)^{5/16} \left( \frac{\Delta t}{10 \yr} \right)^{11/16} \,, \\
	T_{A1}^{\rm N} &= 3.37\times10^{-4}\yr \left( \frac{\Mchirp}{10^{6}M_{\odot}} \right)^{5/8} \left( \frac{\Delta t}{10 \yr} \right)^{3/8} \,.
\end{align}
\end{subequations}
Note that if we \jgb{think of LISA as effectively insensitive below some minimal frequency $f_{\rm min}$, then for sufficiently high $\Mchirp$ this point of entry in the sensitive band will mark the beginning of the signal, obviating the relevance of $f_{\rm start}$}. For $f_{\rm min} = 10^{-5}\Hz$ and $\Delta t \leq 10\yr$, this is the case for $\Mchirp \geq 2.45\times 10^{6}\Msol $. Thus, our final analytical estimates for the $\epsilon$ error measures are built by inserting~\eqref{eq:timescalesNfstart} and~\eqref{eq:estimatederivorb}-\eqref{eq:estimatederivconst} in~\eqref{eq:deffom}, while ensuring the frequency cut $10^{-5}\Hz \leq f \leq 1\Hz$.

\jgb{It is useful to distinguish} between what we will call merging binaries, systems close to coalescence that will merge during the LISA mission lifetime or a few years later, and slowly-chirping binaries still in the deep inspiral phase, of which we observe only a small snapshot in Fourier domain as they do not sweep to the end of the frequency band. The massive black hole binaries (MBH) that will be observed by LISA~\cite{LISA17} fall within the first category, with their merger in band, while the proposed population of stellar-origin black hole binaries (SOBH)~\cite{Sesana16}, with masses comparable to the LIGO/Virgo detections, will comprise both merging binaries, i.e. exiting the LISA band towards larger frequencies during observations, and slowly-chirping binaries hundereds or thousands of years away from merger. Galactic binaries (GB) fall entirely within the second category, and are very close to monochromatic over the lifetime of LISA~\cite{LISA17} \SM{[C: complete]}.

We turn first to the slowly-chirping binaries. For these systems, we will focus on $\epsilon_{\Psi 2}$, which will be the most important error measure. Fig.~\ref{fig:lisafomPsiMcDeltat} shows contour levels for the value of our simple analytical estimate for $\epsilon_{\Psi 2}$ at the beginning of the observations, as a function of the chirp mass and time to merger, for both the orbital and the constellation response. The limit $\epsilon_{\Psi 2} = 1$ is used to single out areas where the perturbative treatment of Sec.~\ref{sec:formalism} is expected to break down, although the precise location of this boundary will vary, depending on the accuracy level required and on the orientation angles. We also single out the high-mass region where the start of the signal is set by its entry \jgb{into the sensitive band, at $f_{\rm min}$ making the error measure independent of time to merger}. We find that, for binaries in the LIGO/Virgo mass range $10^{1}-10^{2} \Msol$, we reach the limit $\epsilon_{\Psi 2}=1$ first for the orbital response, for time-to-merger values within the expected observed range~\cite{Sesana16}. We will introduce in Sec~\ref{subsec:comblisa} an alternative approach to deal with those signals. For completeness, we also show in Fig.~\ref{fig:lisafomPsiMcDeltat} the result for intermediate and massive black hole systems, although we expect to observe mainly merging systems for this mass range.

We now turn to the case of merging binaries. For these we compute the three error measures $\epsilon_{\Psi 2}$, $\epsilon_{A1}$ and $\epsilon_{d}$. To go beyond the crude analytical estimates built from~\eqref{eq:timescalesNfstart} and~\eqref{eq:estimatederivorb}-\eqref{eq:estimatederivconst}, we use numerical derivatives for the timescales~\eqref{eq:defTf}-\eqref{eq:defTA} and for the derivatives in~\eqref{eq:deffom}. We summarize the results in Fig.~\ref{fig:fomLISA}, both for the orbital delay and the constellation response, considering equal-mass systems with total masses of $M=10^{7} \Msol$, $10^{4} \Msol$ and $10^{2} \Msol$ with a starting frequency corresponding to $\Delta t =10 \yr$ of observation. Since individual signals can show significant variation depending on the angular parameters, for the full computation with numerical derivatives we show a geometric average of $\epsilon$ over the sky position, inclination and polarization, together with $\pm 1\sigma$ geometric standard deviation. Here and in the following, for completeness we show results on the frequency band from $10^{-5}\Hz$ to $1\Hz$, while the signals are expected to have little signal-to-noise ratio outside of the band from $10^{-4}\Hz$ to $10^{-1}\Hz$, which is sometimes taken as a reference~\cite{LISA17}.

For $\epsilon_{\Psi 2}$, as expected from~\eqref{eq:TfN} we find a steep rise towards lower frequencies, which shows that considering merging binaries with a limited time to coalescence is crucial here. \jgb{We note differences in} behaviour between the orbital response, where for a fixed mass ratio the initial \jgb{(10yr before merger)} value of $\epsilon_{\Psi 2}$ grows towards lower masses, and the constellation response, where \jgb{initial} $\epsilon_{\Psi 2}$ grows towards higher masses. \jgb{Combining both parts of the response, for equal-mass mergers with $10^2<M<10^7$,} the error measure remains below $\epsilon_{\Psi 2} \sim 0.1$, where the corrections are still well manageable by the perturbative formalism (as we will show in Sec.~\ref{subsec:errorsLISA}).

@@

The $\epsilon_{d}$ is signal-independent, and follows a quite different behaviour for the orbital and constellation responses. For the orbital response, $\epsilon_{d_{0}}$ grows for higher frequencies, reaching the $\sim 0.1$ at around $1\Hz$. For the constellation response, $\epsilon_{d_{c}}$ is mainly important at lower frequencies, reaching $\sim 0.01$ at around $10^{-5}\Hz$.

Overall, we find that for merging binaries the perturbative approach presented in Sec.~\ref{sec:formalism} will be applicable, as the smallness of the annual frequency $f_{0} \simeq 3.1\times10^{-8}\mathrm{Hz}$, compared to the timescales relevant for the signal, ensures a good separation of timescales on the whole LISA frequency band. The most relevant higher-order corrections are expected to be the phase corrections close to the starting frequency of the signal, and the delay correction at high frequencies for the orbital response. For slowly chirping binaries, we have identified regions in the parameter space where the perturbative formalism of Sec.~\ref{sec:formalism} will break down, requiring an alternative approach presented below in Sec.~\ref{subsec:comblisa}.

@@

\SM{[We have to say why we take a non-spinning, equal-mass system - say why the method should also work for precessing signals]}

\SM{[Write explicitly what the orders A1, d1 mean]}

\SM{[argument why we do not compute mismatches]}

\SM{[stress that binaries very far from merger could have arbitrarily large Psi2 corrections]}

\SM{[say that we take a large number of samples to eliminate spline interpolation errors]}

@@

Inserting~\eqref{eq:Gdiscretefourier} into the convolution~\eqref{eq:FDkernel} leads to the following generalized discrete convolution:
\be\label{eq:transferdiscreteconvolution}
	\tilde{s}(f) = \sum_{n \in \mathbb{Z}} c_{n}(f - n f_{0}) \tilde{h} (f - n f_{0}) \,.
\ee
Thus, computing the Fourier-domain response now requires to convolve the signal with a discrete comb with frequency-dependent coefficients $c_{n}$. In practice, this sum is to be truncated at $|n|\leq N$, for some finite order $N$ determining the accuracy of the approximation. To assess the expected truncation error, we use a simple criterion based on the $L^{1}$-norm of the $\{c_{n}\}$ sequence. For a given target truncation error $\eta$, we define the truncation \SM{change notation to avoid confusion with $N$ from Fresnel stencil} $N(f, \eta)$ as the smallest integer such that 
\be\label{eq:criteriontruncationcomb}
	\sum_{|n| > N(f, \eta)} |c_{n}(f)| < \eta \,.
\ee
The truncation order $N(f,\eta)$ is frequency-dependent, and also depends on the orientation angles. For $c_{n}^{L}$, it also determines the number of function evaluations and the cost of the FFT. For $c_{n}^{0}$, the expression~\eqref{eq:cn0} can provide an asymptotic bound on the width of the comb. For large values of $n$ we have indeed the equivalent $| J_{n}[z] | \sim (e z/2n)^{n}/\sqrt{2\pi n} \; \left[ n\rightarrow + \infty \right]$~\cite{} \SM{[C: dlmf?]}, so that a conservative estimate for the truncation order (almost independent of $\eta$) is given by $N \geq  |e \pi f R \cos \beta|$. Fig.~\ref{fig:lisacombextent} shows numerical computations for $N(f,\eta)$, averaged over orientation angles, for both the constellation and the orbital response and for the truncation levels  $\eta=10^{-12}$, $10^{-6}$, $10^{-3}$. We see that, especially at high frequencies, the orbital response requires more coefficients than the constellation response due to the longer baseline of the delay. At the frequency $f=0.1\Hz$, we have for the orbital response $N \simeq 200$.

@@

\begin{figure}
  \centering
%  \includegraphics[width=.45\linewidth]{plots/eulerSXS0058alphagamma.pdf}
%  \hspace{0.2cm}
%  \includegraphics[width=.45\linewidth]{plots/eulerSXS0058beta.pdf}
  \includegraphics[width=.98\linewidth]{plots/eulerSXS0058_py.pdf}
  \caption{Evolution of Euler angles $(\alpha, \beta, \gamma)$ near merger for the example SpEC waveform SXS:BBH:0058~\cite{SpEC, SXScatalog, Mroue+12, Mroue+13} \SM{[C]}. Full, dashed and dotted lines represent the NR, SEOB and extended SEOB waveforms respectively. The vertical line at $t=0$ indicates the time of merger. In the left panel the full and dashed black lines indicate the asymptotic behaviour~\eqref{eq:OmegaframeQNM} for the rotation of the frame around the direction of the final $\bm{J}$. The difference between NR and SEOB there is due to the value of the final spin, with $\chi_{f}^{\rm NR} = 0.54$ and $\chi_{f}^{\rm SEOB} = 0.42$, which leads to a different $\Omega_{\rm frame}$. In the right panel, $\beta$ is asymptotically constant in our toy model. \SM{[Add amplitude/$\omega$ plots comparing NR/SEOB, and possibly PhenomD for the 22 mode ? Add Euler angles produced by IFFT of PhenomP ?]} \SM{[remove SEOBext in the legend]} \SM{[confusion of dashed black lines]}}
  \label{fig:precmodel}
\end{figure}

@@

\SM{[add plot of QNM frequency and frequency diff according to final spin ?]}.

@@

The results are shown in Fig.~\ref{fig:fomprec}, using the merger frequency and the ringdown frequency, shown by the vertical lines, to give an idea of the separation between the inspiral and post-merger phases. We can take $\epsilon \sim 1$ as an order-of-magnitude indication of the breakdown of a perturbative treatment. An important point about Fig.~\ref{fig:fomprec} is that the error estimates are relative for each mode, thus higher $|m-m'|$ modes, which are found to be the hardest to model precisely, can be very subdominant in the final waveform.

The magnitude of the error estimates for $++$ and $\perp\perp$ is roughly in agreement with the PN-inspired computation~\eqref{eq:precfomPN} above for the inspiral part of the signal \SM{[check if the PN is quantitatively ok -- overlay it on the plot ?]} \SM{[Pb: scaling of FOM might be dominated here by SS oscillations, notably for 2nd der. ? To check - check also geometric scaling from mode to mode]}, with $\epsilon_{\Psi 2}$ showing a negative slope $v^{-1}=(Mf)^{-1/3}$, and with a hierarchy between modes due to the factors of $|m' \cos \beta - m|$. The case $--$ departs from the simple precession picture, and~\eqref{eq:precfomPN} does not apply. Before merger, we see that amplitude corrections are within the preturbative regime, contrarily to $\epsilon_{\Psi 2}$ that can exceed $1$ for subdominant modes in the $++$ and $\perp\perp$ cases. The amplitude corrections are found to be in the perturbative regime for all cases during the inspiral, however both the $++$ and $\perp\perp$ cases show a sharp increase of $\epsilon_{A1}$, $\epsilon_{A2}$ post-merger. The case $--$, with its small remnant spin and mild post-merger frame rotation, shows no such increase.

@@

\SM{[possibly move this to the formalism section ? but plots for our three examples]}

\SM{[say that another possibility would be to go back to time-domain and FFT for the limited-length near-merger part of the signal]}

\SM{[expand motivation: sharp feature TD $\rightarrow$ smooth in FD $\rightarrow$ hope to have a cheap trigo representation of FD merger range]}

@@

\jgb{This work was supported by a NASA Astrophysics Theory Program grant 11-ATP-046.}

>>>>

The \jgb{constellation} response~\eqref{eq:GslrL}-\eqref{eq:transferLlocal} can be interpreted as the Fourier-domain translation of a discrete derivative taken on the waveform. The \jgb{leading factor in~\eqref{eq:GslrL}} shows that the amplitude of the response is proportional to $f$ in the low-frequency limit $f\ll f_{L}$, where the other factors are essentially unity. For $f\gtrsim f_{L}$, the $\sinc$ and the phase of the exponential generate additional structure in the response, including zero-crossings when the projected armlength \jgb{is an integer number of wavelengths}. From~\eqref{eq:transferLlocal}, an expansion for small $f\ll f_{L}$ yields back a Fourier-domain analog of the low-frequency approximation of the response~\cite{Cutler97, RCP04}, which \jgb{is equivalent to} having two LIGO-type interferometers turned by $\pi/4$ and set in motion.

For analysis of the response we need a concrete set of gravitational waveforms. We will use the PhenomD model~\cite{Khan+15,Husa+15}, which provides Fourier-domain inspiral-merger-ringdown waveforms for aligned spins.

@@

To obtain an order-of-magnitude estimate for the error measures $\epsilon$~\eqref{eq:deffom}, we will use the Newtonian-order expressions~\eqref{eq:timescalesN} for the signal-dependent timescales $\Tf$ and $T_{A1}$. Higher-order amplitude terms beyond the first one in~\eqref{eq:resulttaylorA}, as well as phase terms beyond the second derivative in~\eqref{eq:expandPsi}, will turn out to be  negligible in the LISA case and we will ignore them in the following. It is useful to separate the orbital response~\eqref{eq:transfer0local} and the constellation response~\eqref{eq:transferLlocal}, as the different baseline of the delays (orbital radius $R$ or armlength $L$) as well as the presence of a time-varying prefactor $F(t)$ both affect the result. When estimating the magnitude of the relevant derivatives of $G$, we must also take into account dimensionless delay factors of the type $2\pi f d$.

We start with the orbital response, which takes the form of a pure delay $G_{0}(f, t) = e^{-2 i \pi f d_{0}(t)}$, and obtain
\begin{subequations}
\begin{align}
	\frac{1}{G_{0}} \partial_{t} G_{0} &= -2i\pi f \dot{d}_{0}\,,\\
	\frac{1}{G_{0}} \partial_{tt} G_{0} &= -2i\pi f \ddot{d}_{0} - 4\pi^{2} f^{2} \dot{d}_{0}^{2} \,,\\
	\frac{1}{G_{0}} \partial_{tf} G_{0} &= -2 i \pi \dot{d}_{0} - 4\pi^{2} f d_{0} \dot{d}_{0} \,.
\end{align}
\end{subequations}

Since the one-arm constellation response~\eqref{eq:decomposeGslr}-\eqref{eq:transferLlocal} is analogous to a discrete time derivative of the signal, it is appropriate to keep explicit an overall factor $f$ reflecting this structure. Hence we write symbolically $G_{L}(f,t) \sim f F(t) e^{-2i \pi f d_{L}(t)}$, where $d_{L}$ represents a delay term and $F(t)$ represents the rest of the geometric factors in~\eqref{eq:transferLlocal}, for which we momentarily ignore the $f$-dependence. This gives
\begin{subequations}
\begin{align}
	\frac{1}{G_{L}} \partial_{t} G_{L} &\sim -2i\pi f \dot{d}_{c} + \frac{\dot{F}}{F}\,,\\
	\frac{1}{G_{L}} \partial_{tt} G_{L} &\sim -2i\pi f \ddot{d}_{c} - 4\pi^{2} f^{2} \dot{d}_{c}^{2} - 2i\pi f \dot{d}_{c} \frac{\dot{F}}{F} - \frac{\dot{F}^{2}}{F^{2}} + \frac{\ddot{F}}{F} \,,\\
	\frac{1}{G_{L}} \partial_{tf} G_{L} &\sim -4 i \pi \dot{d}_{c} - 2i\pi d_{c} \frac{\dot{F}}{F} - 4\pi^{2} f d_{c} \dot{d}_{c} + \frac{1}{f}\frac{\dot{F}}{F}\,.
\end{align}
\end{subequations}

@@

An important point in~\eqref{eq:estimatederivorb} and~\eqref{eq:estimatederivconst} above is the presence of delay factors in the forms of powers of $2\pi f d$, so that we cannot use the simple replacement $\partial_{t}^{n}G \rightarrow \Omega_{0}^{n}G$. Thus, the suitability of the leading-order treatment cannot be estimated by a mere separation between the signal timescales and the annual timescale for the response, but will also depend on the frequency being above or below the transfer frequencies $f_{R}$ and $f_{L}$ defined in~\eqref{eq:transferfrequencies}. The error measure $\epsilon_{d}$ does not depend on signal-dependent timescales and can be directly read off the estimates above, with $\epsilon_{d} = |\partial_{tf}G/(2\pi G)|$. For the error measures $\epsilon_{\Psi 2}$ and $\epsilon_{A1}$, we will combine the above derivatives with the Newtonian timescales given in~\eqref{eq:timescalesN} for the inspiral phase of the signal.

The starting frequency of the signal will play an important role. As a function of the time remaining before merger $\Delta t$, from the Newtonian relation~\eqref{eq:omegaphiN} we have (using the chirp mass $\Mchirp = M \nu^{3/5}$):
\be\label{eq:fstartN}
	f_{\rm start} = 1.75\times 10^{-5} \Hz \left( \frac{\Mchirp}{10^{6}M_{\odot}} \right)^{-5/8} \left( \frac{\Delta t}{10 \yr} \right)^{-3/8} \,.
\ee
This gives in turn for the Newtonian timescales~\eqref{eq:timescalesN}
\begin{subequations}\label{eq:timescalesNfstart}
\begin{align}
	\Tf^{\rm N} &= 8.78\times10^{-2}\yr \left( \frac{\Mchirp}{10^{6}M_{\odot}} \right)^{5/16} \left( \frac{\Delta t}{10 \yr} \right)^{11/16} \,, \\
	T_{A1}^{\rm N} &= 3.37\times10^{-4}\yr \left( \frac{\Mchirp}{10^{6}M_{\odot}} \right)^{5/8} \left( \frac{\Delta t}{10 \yr} \right)^{3/8} \,.
\end{align}
\end{subequations}
Note that if we think of LISA as effectively insensitive below some minimal frequency $f_{\rm min}$, then for sufficiently high $\Mchirp$ this point of entry in the sensitive band will mark the beginning of the signal, obviating the relevance of $f_{\rm start}$. For $f_{\rm min} = 10^{-5}\Hz$ and $\Delta t \leq 10\yr$, this is the case for $\Mchirp \geq 2.45\times 10^{6}\Msol $. Thus, our final analytical estimates for the $\epsilon$ error measures are built by inserting~\eqref{eq:timescalesNfstart} and~\eqref{eq:estimatederivorb}-\eqref{eq:estimatederivconst} in~\eqref{eq:deffom}, while ensuring the frequency cut $10^{-5}\Hz \leq f \leq 1\Hz$.

It is useful to distinguish between what we will call merging binaries, systems close to coalescence that will merge during the LISA mission lifetime or a few years later, and slowly-chirping binaries still in the deep inspiral phase, of which we observe only a small snapshot in Fourier domain as they do not sweep to the end of the frequency band. The massive black hole binaries (MBH) that will be observed by LISA~\cite{LISA17} fall within the first category, with their merger in band, while the proposed population of stellar-origin black hole binaries (SOBH)~\cite{Sesana16}, with masses comparable to the LIGO/Virgo detections, will comprise both merging binaries, i.e. exiting the LISA band towards larger frequencies during observations, and slowly-chirping binaries hundereds or thousands of years away from merger.

We turn first to the slowly-chirping binaries. For these systems, we will focus on $\epsilon_{\Psi 2}$, which will be the most important error measure. Fig.~\ref{fig:lisafomPsiMcDeltat} shows contour levels for the value of our simple analytical estimate for $\epsilon_{\Psi 2}$ at the beginning of the observations, as a function of the chirp mass and time to merger, for both the orbital and the constellation response. The limit $\epsilon_{\Psi 2} = 1$ is used to single out areas where the perturbative treatment of Sec.~\ref{sec:formalism} is expected to break down, although the precise location of this boundary will vary, depending on the accuracy level required and on the orientation angles. We also single out the high-mass region where the start of the signal is set by its entry into the sensitive band at $f_{\rm min}$, making the error measure independent of the time to merger. We find that, for binaries in the LIGO/Virgo mass range $10^{1}-10^{2} \Msol$, we reach the limit $\epsilon_{\Psi 2}=1$ first for the orbital response, for time-to-merger values within the expected observed range~\cite{Sesana16}. We will introduce in Sec~\ref{subsec:comblisa} an alternative approach to deal with those signals. For completeness, we also show in Fig.~\ref{fig:lisafomPsiMcDeltat} the result for intermediate and massive black hole systems, although we expect to observe merging systems for this mass range.

We now turn to the case of merging binaries. For these we compute the three error measures $\epsilon_{\Psi 2}$, $\epsilon_{A1}$ and $\epsilon_{d}$. To go beyond the crude analytical estimates built from~\eqref{eq:timescalesNfstart} and~\eqref{eq:estimatederivorb}-\eqref{eq:estimatederivconst}, we use numerical derivatives for the timescales~\eqref{eq:defTf}-\eqref{eq:defTA} and for the derivatives in~\eqref{eq:deffom}. We summarize the results in Fig.~\ref{fig:fomLISA}, both for the orbital delay and the constellation response, considering equal-mass non-spinning systems with total masses of $M=10^{7} \Msol$, $10^{4} \Msol$ and $10^{2} \Msol$ with a starting frequency corresponding to $\Delta t =10 \yr$ of observation. Since individual signals can show significant variation depending on the orientation parameters, for the full computation with numerical derivatives we show a geometric average of $\epsilon$ over the sky position, inclination and polarization, together with $\pm 1\sigma$ geometric standard deviation. Here and in the following, for completeness we show results on the frequency band from $10^{-5}\Hz$ to $1\Hz$, while the signals are expected to have little signal-to-noise ratio outside of the band from $10^{-4}\Hz$ to $10^{-1}\Hz$, which is sometimes taken as a reference~\cite{LISA17}.

For $\epsilon_{\Psi 2}$, as expected from~\eqref{eq:TfN} we find a steep rise towards lower frequencies, which shows that considering merging binaries with a limited time to coalescence is crucial here. We note differences in behaviour between the orbital response, where for a fixed mass ratio the initial (10yr before merger) value of $\epsilon_{\Psi 2}$ grows towards lower masses, and the constellation response, where the initial $\epsilon_{\Psi 2}$ grows towards higher masses. Combining both parts of the response, for mergers with $10^2<M<10^7$, the error measure remains below $\epsilon_{\Psi 2} \sim 0.1$, where the corrections are still well manageable by the perturbative formalism (as we will show in Sec.~\ref{subsec:errorsLISA}).

@@

The $\epsilon_{d}$ is signal-independent, and follows a quite different behaviour for the orbital and constellation responses. For the orbital response, $\epsilon_{d_{0}}$ grows for higher frequencies, reaching the $\sim 0.1$ at around $1\Hz$. For the constellation response, $\epsilon_{d_{L}}$ is mainly important at lower frequencies, reaching $\sim 0.01$ at around $10^{-5}\Hz$.

Overall, we find that for merging binaries the perturbative approach presented in Sec.~\ref{sec:formalism} will be applicable on the whole LISA frequency band. The most relevant higher-order corrections are expected to be the phase corrections close to the starting frequency of the signal, and the delay correction at high frequencies for the orbital response. For slowly chirping binaries, we have identified regions in the parameter space where the perturbative formalism of Sec.~\ref{sec:formalism} will break down, requiring an alternative approach presented below in Sec.~\ref{subsec:comblisa}.

@@

@@

Inserting~\eqref{eq:Gdiscretefourier} into the convolution~\eqref{eq:FDkernel} leads to the following generalized discrete convolution:
\be\label{eq:transferdiscreteconvolution}
	\tilde{s}(f) = \sum_{n \in \mathbb{Z}} c_{n}(f - n f_{0}) \tilde{h} (f - n f_{0}) \,.
\ee
Thus, computing the Fourier-domain response now requires to convolve the signal with a discrete comb with frequency-dependent coefficients $c_{n}$. In practice, this sum is to be truncated at $|n|\leq N$, for some finite order $N$ determining the accuracy of the approximation. To assess the expected truncation error, we use a simple criterion based on the $L^{1}$-norm of the $\{c_{n}\}$ sequence. For a given target truncation error $\eta$, we define the truncation \SM{change notation to avoid confusion with $N$ from Fresnel stencil} $N(f, \eta)$ as the smallest integer such that 
\be\label{eq:criteriontruncationcomb}
	\sum_{|n| > N(f, \eta)} |c_{n}(f)| < \eta \,.
\ee
The truncation order $N(f,\eta)$ is frequency-dependent, and also depends on the orientation angles. For $c_{n}^{L}$, it also determines the number of function evaluations and the cost of the FFT. For $c_{n}^{0}$, the expression~\eqref{eq:cn0} can provide an asymptotic bound on the width of the comb. For large values of $n$ we have indeed the equivalent $| J_{n}[z] | \sim (e z/2n)^{n}/\sqrt{2\pi n} \; \left[ n\rightarrow + \infty \right]$~\cite{DLMF} \SM{[C: dlmf?]}, so that a conservative estimate for the truncation order (almost independent of $\eta$) is given by $N \geq  |e \pi f R \cos \beta|$. Fig.~\ref{fig:lisacombextent} shows numerical computations for $N(f,\eta)$, averaged over orientation angles, for both the constellation and the orbital response and for the truncation levels  $\eta=10^{-12}$, $10^{-6}$, $10^{-3}$. We see that, especially at high frequencies, the orbital response requires more coefficients than the constellation response due to the longer baseline of the delay. At the frequency $f=0.1\Hz$, we have for the orbital response $N \simeq 200$.

@@

\begin{figure}
  \centering
%  \includegraphics[width=.45\linewidth]{plots/eulerSXS0058alphagamma.pdf}
%  \hspace{0.2cm}
%  \includegraphics[width=.45\linewidth]{plots/eulerSXS0058beta.pdf}
  \includegraphics[width=.98\linewidth]{plots/eulerSXS0058_py.pdf}
  \caption{Evolution of Euler angles $(\alpha, \gamma)$ (top panel) and $\beta$ (bottom panel) near merger for the example SpEC waveform SXS:BBH:0058~\cite{SpEC, SXScatalog, Mroue+12, Mroue+13}. Full, dashed and dotted lines represent the NR, SEOB and extended SEOB waveforms respectively. The vertical line at $t=0$ indicates the time of merger. In the left panel the full and dashed black lines indicate the asymptotic behaviour~\eqref{eq:OmegaframeQNM} for the rotation of the frame around the direction of the final $\bm{J}$. The difference between NR and SEOB there is due to the value of the final spin, with $\chi_{f}^{\rm NR} = 0.54$ and $\chi_{f}^{\rm SEOB} = 0.42$, which leads to a different $\Omega_{\rm frame}$. In the right panel, $\beta$ is asymptotically constant in our toy model. \SM{[Add amplitude/$\omega$ plots comparing NR/SEOB, and possibly PhenomD for the 22 mode ? Add Euler angles produced by IFFT of PhenomP ?]}}
  \label{fig:precmodel}
\end{figure}

@@

@@

The results are shown in Fig.~\ref{fig:fomprec}, using the merger frequency and the ringdown frequency, shown by the vertical lines, to give an idea of the separation between the inspiral and post-merger phases. The leading-order PN prediction for single-spin simple precession~\eqref{eq:precfomPN} is overlayed for the cases $++$ and $\perp\perp$, but not for the case $--$ as it departs significantly from simple precession. We can take $\epsilon \sim 1$ as an order-of-magnitude indication of the breakdown of a perturbative treatment. An important point about Fig.~\ref{fig:fomprec} is that the error estimates are relative for each mode, thus higher $|m-m'|$ modes, which are found to be the hardest to model precisely, can be very subdominant in the final waveform.

The magnitude of the error estimates for $++$ and $\perp\perp$ is roughly in agreement with the PN-inspired computation~\eqref{eq:precfomPN} above for the inspiral part of the signal, with $\epsilon_{\Psi 2}$ showing a negative slope $v^{-1}=(Mf)^{-1/3}$, and with a hierarchy between modes due to the factors of $|m' \cos \beta - m|$. The PN estimates are missing oscillations on the precession timescale due to the double-spin precession. The case $--$ departs from the simple precession picture, and~\eqref{eq:precfomPN} does not apply. Before merger, we see that amplitude corrections are within the preturbative regime, contrarily to $\epsilon_{\Psi 2}$ that can exceed $1$ for subdominant modes in the $++$ and $\perp\perp$ cases. The amplitude corrections are found to be in the perturbative regime for all cases during the inspiral, however both the $++$ and $\perp\perp$ cases show a sharp increase of $\epsilon_{A1}$, $\epsilon_{A2}$ post-merger. The case $--$, with its small remnant spin and mild post-merger frame rotation, shows no such increase.

@@

@@

We are indebted to Alejandro Boh\'{e}, Alessandra Buonanno, Scott Field, Philip Graff, Sebastian Khan, Antoine Klein, Serguei Ossokine, Michael P\"{u}rrer for useful discussions and comments. We used the software GWFrames~\cite{Boyle+14} in our study. This work was supported by a NASA Astrophysics Theory Program grant 11-ATP-046.
